---
title: "Ebola"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE}
library(dplyr)
library(stringr)
library(lubridate)
```

```{r}
library(outbreaks)
data("ebola_sim")

```


#Capitalize first letter for male and female

#Remove white space from hospital column

#Create a delay_days column that defines difference between report_date and onset_date

#Convert outcome column to lowercase, replace NA's with the word unknown

#Convert dates to ymd format
```{r}


cleaned_ebola <- ebola_sim$linelist %>%
  mutate(
    gender = str_to_title(str_trim(gender)),
    hospital = str_squish(hospital),
    outcome = str_to_lower(str_replace_na(outcome, "unknown")),
    onset_date = ymd(date_of_onset),
    report_date = ymd(date_of_hospitalisation), 
    delay_days = as.numeric(report_date - onset_date)
  ) %>%
  filter(!is.na(onset_date))


```


#Extract columns and show example of cbind function

```{r}

#cbind example
dates <- as.character(cleaned_ebola$date_of_outcome)
cases <- cleaned_ebola$outcome
hospital <- cleaned_ebola$hospital
cbind_ex <- cbind(dates, hospital, cases)
head(data.frame(cbind_ex))
```

#Create a quarter column that shows the quarter of the year that date of hospitalsation occurred

```{r}
library(dplyr)
library(lubridate)

cleaned_ebola <- ebola_sim$linelist %>%
  mutate(
    date_of_hospitalisation = ymd(date_of_hospitalisation),
    quarter_of_hospitalisation = quarter(date_of_hospitalisation, with_year = FALSE, fiscal_start = 1)
  )

```

#Show list of distinct hospitals

```{r}

unique_hospitals <- cleaned_ebola %>%
  filter(!is.na(hospital)) %>%
  distinct(hospital)


print(unique_hospitals)

```
#Create case when statement to clean incorrectly typed hospital names

```{r}
library(dplyr)
library(stringr)

cleaned_ebola <- cleaned_ebola %>%
  mutate(
    hospital = case_when(
      str_detect(hospital, "Military Hospital") ~ "Military Hospital",
      str_detect(hospital, "Connaught Hospital") ~ "Connaught Hospital",
      str_detect(hospital, "Princess Christian Maternity Hospital") ~ "Princess Christian Maternity Hospital (PCMH)",
      str_detect(hospital, "Rokupa Hospital") ~ "Rokupa Hospital",
      str_detect(hospital, "Mitylira Hospital") ~ "Mitylira Hospital",
      str_detect(hospital, "Rokupa Hopital") ~ "Rokupa Hospital",
      str_detect(hospital, "Princess Christian Maternity Hopital") ~ "Princess Christian Maternity Hospital (PCMH)",
      str_detect(hospital, "Connaught Hopital") ~ "Connaught Hospital",
      str_detect(hospital, "Military Hopital") ~ "Military Hospital",
      str_detect(hospital, "Mitylira Hopital") ~ "Mitylira Hospital",
      TRUE ~ hospital  # Keep original name if no match is found
    )
  )

```


```{r}


case_counts <- cleaned_ebola %>%
  filter(
    !is.na(hospital),
    !is.na(quarter_of_hospitalisation),
    year(date_of_hospitalisation) == 2014
  ) %>%
  group_by(hospital, quarter_of_hospitalisation) %>%
  summarise(n_cases = n()) %>%
  ungroup()

```


#Create a graph to count number of cases by quarter in 2024 and color coordinate by hospital name
```{r}
library(ggplot2)

ggplot(case_counts, aes(x = quarter_of_hospitalisation, y = n_cases, fill = hospital)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Ebola Cases by Hospital and Quarter (2014)",
    x = "Quarter of Hospitalisation",
    y = "Number of Cases",
    fill = "Hospital"
  ) +
  theme_minimal()

```

#Use pivot_wider function to make a table that shows the total number of hospitilisations per year by hospital name

```{r}
library(dplyr)
library(tidyr)
library(lubridate) 

hospital_case_counts <- cleaned_ebola %>%
  mutate(year_of_hospitalisation = year(date_of_hospitalisation)) %>%  
  group_by(hospital, year_of_hospitalisation) %>%
  summarise(n_cases = n(), .groups = "drop")


hospital_case_counts_wide <- hospital_case_counts %>%
  pivot_wider(
    names_from = hospital,
    values_from = n_cases,
    values_fill = list(n_cases = 0)
  )

hospital_case_counts_wide


```
#List number of NA's per column

```{r}

sapply(cleaned_ebola, function(x) sum(is.na(x)))

```


```{r}
lapply(cleaned_ebola, summary)

```

```{r}

survived_ebola <- subset(cleaned_ebola, outcome == "Recover")

head(survived_ebola)


```

```{r}


deaths_data <- cleaned_ebola %>%
  filter(outcome == "Death" & !is.na(date_of_outcome)) %>% 
  mutate(year = year(ymd(date_of_outcome))) %>%
  group_by(year) %>%
  summarise(deaths = n())  


total_cases_by_year <- cleaned_ebola %>%
  filter(!is.na(date_of_outcome)) %>%  
  mutate(year = year(ymd(date_of_outcome))) %>%
  group_by(year) %>%
  summarise(total_cases = n()) 


death_data <- left_join(deaths_data, total_cases_by_year, by = "year")


death_data <- death_data %>%
  mutate(death_rate = deaths / total_cases)


ggplot(death_data, aes(x = factor(year), y = death_rate, fill = "red")) +
  geom_bar(stat = "identity") +
  labs(
    title = "Death Rate by Year",
    x = "Year",
    y = "Death Rate (Proportion)"
  ) +
  theme_minimal()


print(deaths_data)


print(death_data)

```

```{r}



cleaned_ebola <- cleaned_ebola %>%
  mutate(year = year(ymd(date_of_outcome))) %>%
  filter(!is.na(date_of_outcome)) 


cases_per_year <- table(cleaned_ebola$year)


deaths_data <- cleaned_ebola %>%
  filter(outcome == "Death")


deaths_per_year <- table(deaths_data$year)


death_rate <- deaths_per_year / cases_per_year


cat("Deaths by Year:\n")
print(deaths_per_year)

cat("\nDeath Rate by Year:\n")
print(death_rate)



mean_death_rate <- mean(death_rate, na.rm = TRUE)
median_death_rate <- median(death_rate, na.rm = TRUE)
sd_death_rate <- sd(death_rate, na.rm = TRUE)
min_death_rate <- min(death_rate, na.rm = TRUE)
max_death_rate <- max(death_rate, na.rm = TRUE)

cat("\nSummary Statistics for Death Rate:\n")
cat("Mean death rate:", mean_death_rate, "\n")
cat("Median death rate:", median_death_rate, "\n")
cat("Standard deviation:", sd_death_rate, "\n")
cat("Min death rate:", min_death_rate, "\n")
cat("Max death rate:", max_death_rate, "\n")

```

```{r}

valid_rows <- !is.na(cleaned_ebola$date_of_outcome)
data_valid <- cleaned_ebola[valid_rows, ]

data_valid$year_outcome <- as.numeric(format(as.Date(data_valid$date_of_outcome), "%Y"))


death_counts <- table(data_valid$year_outcome[data_valid$outcome == "Death"])


total_counts <- table(data_valid$year_outcome)


death_rate <- death_counts / total_counts


print(death_counts)
print(death_rate)

```

```{r}

ebola_sub <- cleaned_ebola[!is.na(cleaned_ebola$outcome) & !is.na(cleaned_ebola$gender), ]


ebola_sub$date_of_outcome <- as.Date(ebola_sub$date_of_outcome)
ebola_sub$year <- format(ebola_sub$date_of_outcome, "%Y")


gender_outcome_year <- table(ebola_sub$gender, ebola_sub$outcome, ebola_sub$year)


print(gender_outcome_year)


total_cases <- apply(gender_outcome_year, c(1, 3), sum)

deaths_only <- gender_outcome_year[, "Death", ]


death_rate <- deaths_only / total_cases


cat("Death rates by gender and year:\n")
print(round(death_rate, 3))

```

```{r}



bar_locs <- barplot(death_rate,
                    beside = TRUE,
                    col = c("lightblue", "pink"),
                    main = "Death Rate by Gender and Year",
                    xlab = "Year",
                    ylab = "Death Rate",
                    ylim = c(0, 1),
                    names.arg = colnames(death_rate))


legend("topright",
       legend = c("Male", "Female"),
       fill = c("lightblue", "pink"),
       title = "Gender")
```

