---
title: "Influenza"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Use for loops to read through multiple datasets

#Reduce datasets to only share common columns

#Rename each dataset

```{r}
for (i in 2009:2014) {
  file_name <- paste0("Influenza_", i, ".csv")
  flu_data <- read.csv(file_name)
  var_name <- paste0("data", i)

  assign(var_name, flu_data)
}

c_columns <- Reduce(intersect, lapply(2009:2014, function(i) colnames(get(paste0("data", i)))) )

for (i in 2009:2014) {
  var_name <- paste0("data", i) 
  current_data <- get(var_name)
  current_data <- current_data[, c_columns, drop = FALSE]
  assign(var_name, current_data)
}



```

#Add year column to each dataset

#Rbind data from 2009-2014

```{r}

years <- 2009:2014
final_data <- NULL 

for (year in years) {
  dataset_name <- paste0("data", year)
  dataset <- get(dataset_name)
  dataset$year <- year  
  final_data <- rbind(final_data, dataset)
}
head(final_data)
```

```{r,include=FALSE}
library(dplyr)
library(ggplot2)
```

#Plot data grouped by region and year

```{r}
summary_d <- final_data %>%
  group_by(Region, year) %>% # Summarize combined_data by Region and Year
  summarize(
    total_patient_visits = sum(Total_Patient_Visits, na.rm = TRUE),  
    total_ili = sum(Total_ILI, na.rm = TRUE)                        
  )

# Plot the data with adjustments
ggplot(summary_d, aes(x = total_patient_visits, y = total_ili, color = Region)) + 
  geom_point(aes(size = total_ili), alpha = 0.7) +  
  scale_x_continuous(trans = 'log10') +  
  scale_y_continuous(trans = 'log10') +  
  theme_minimal() + 
  labs(
    title = "Influenza Total Visits vs Total ILI",
    x = "Total Patient Visits (log scale)",
    y = "Total ILI (log scale)"
  )
```

#Filter for weeks 35 - 45 in year 2009

```{r}


#combined_data <- combined_data %>%
#  mutate(week_year = as.numeric(paste0(year, #sprintf("%02d", Week))))
data_selected_weeks <- final_data %>% 
  filter(Week >= 200935 & Week <= 200945)

week_plot <- data_selected_weeks %>%
  group_by(Week) %>%
  summarize(
    total_patient_visits = sum(Total_Patient_Visits, na.rm = TRUE),
    total_ili = sum(Total_ILI, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(week_plot, aes(x = Week, y = total_patient_visits, color = as.factor(Week))) + 
  geom_line() +  
  geom_point(size = 3) + 
  theme_minimal() +
  labs(
    title = "Total Patient Visits for Influenza from August 24, 2009 to November 8, 2009",
    x = "Week-Year (2009)",
    y = "Total Patient Visits"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    legend.position = "none"  
  )

```

#Historgram of 2010 Weighted ILI Data

```{r}




data_2010 <- final_data %>%
  filter(year == 2010)


ggplot(data_2010, aes(x = Perc_Weighted_ILI)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  theme_minimal() +
  labs(
    title = "Distribution of Weighted ILI (%) in 2010",
    x = "Weighted ILI (%)",
    y = "Frequency"
  )


```



#Apply Sapply and cbind functions

```{r}
numeric_columns <- final_data[, 6:9]
means <- sapply(numeric_columns, mean, na.rm = TRUE)


SDs <- sapply(numeric_columns, sd, na.rm = TRUE)


results_table <- cbind(means, SDs)

results_table

```
#Show calculations by

# a) method of using summarize 

# b) method of using tapply function

```{r}
library(dplyr)

summ <- final_data %>%
  group_by(year) %>%
  summarize(mean_visits = mean(Total_Patient_Visits, na.rm = TRUE))

print(summ)

```
```{r}
tapply(final_data$Total_Patient_Visits, final_data$year, mean, na.rm = TRUE)

```

